
---
title: "Lexical Diversity & Sentiment in Amazon Fine Food Reviews"
author:
  - David Nguyen
  - Ayushi Mazumder
  - Charles Martin
  - Ryan Kuhn
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
    code_folding: hide
---

```{r setup, include=FALSE}
# Prevent CRAN prompts
options(repos = c(CRAN = "https://cloud.r-project.org"))
knitr::opts_chunk$set(
  echo        = TRUE,
  message     = FALSE,
  warning     = FALSE,
  fig.width   = 8,
  fig.height  = 5,
  fig.align   = "center"
)
````

## Introduction

The Amazon Fine Food Reviews dataset contains over 500,000 user reviews for food products sold on Amazon.
This analysis addresses two key questions:

1. **Lexical Diversity**
   Do negative reviews (1–2 stars) exhibit lower lexical diversity than positive reviews (4–5 stars)?
2. **Sentiment Intensity**
   Do negative reviews contain a higher proportion of negative words than positive reviews?

We measure lexical diversity using the Type-Token Ratio (TTR) and negative sentiment intensity as the proportion of words found in the “negative” category of the Bing sentiment lexicon.

---

## 1. Install & Load Packages

```{r packages}
required_packages <- c("tidyverse", "tidytext", "sentimentr", "lubridate")
installed <- rownames(installed.packages())
for (pkg in required_packages) {
  if (! pkg %in% installed) {
    install.packages(pkg, dependencies = TRUE)
  }
}
library(tidyverse)
library(tidytext)
library(sentimentr)
library(lubridate)
```

---

## 2. Load Data

```{r load-data}
# Assumes Reviews.csv is in data/ subfolder
reviews_raw <- read_csv("data/Reviews.csv")
glimpse(reviews_raw)
```

---

## 3. Preprocessing

* Convert Unix timestamp to POSIX date.
* Cast Score to numeric.
* Create a `Sentiment` factor: Negative (1–2), Positive (4–5), exclude 3-star.

```{r preprocess}
reviews <- reviews_raw %>%
  mutate(
    Date      = as_datetime(Time),
    Score     = as.numeric(Score),
    Sentiment = case_when(
      Score %in% c(1,2) ~ "Negative",
      Score %in% c(4,5) ~ "Positive",
      TRUE              ~ NA_character_
    )
  ) %>%
  filter(!is.na(Sentiment)) %>%
  select(Id, ProductId, UserId, ProfileName,
         HelpfulnessNumerator, HelpfulnessDenominator,
         Score, Date, Summary, Text, Sentiment)
```

---

## 4. Lexical Diversity (TTR)

Define and compute TTR for each review:

```{r compute-ttr}
compute_ttr <- function(text) {
  tokens <- unlist(str_split(text, "\\s+"))
  tokens <- tokens[tokens != ""]
  if (length(tokens) == 0) return(NA_real_)
  length(unique(tokens)) / length(tokens)
}

reviews <- reviews %>%
  mutate(TTR = map_dbl(Text, compute_ttr))
```

---

## 5. Negative Sentiment Intensity

Count negative words per review and calculate intensity:

```{r sentiment-intensity}
negative_lexicon <- get_sentiments("bing") %>%
  filter(sentiment == "negative")

neg_counts <- reviews %>%
  unnest_tokens(word, Text) %>%
  inner_join(negative_lexicon, by = "word") %>%
  count(Id, name = "NegativeCount")

reviews <- reviews %>%
  left_join(neg_counts, by = "Id") %>%
  mutate(
    NegativeCount     = replace_na(NegativeCount, 0),
    TotalWords        = str_count(Text, "\\S+"),
    NegativeIntensity = NegativeCount / TotalWords
  )
```

---

## 6. Summary Statistics

```{r summary-stats}
summary_stats <- reviews %>%
  group_by(Sentiment) %>%
  summarise(
    Count            = n(),
    Mean_TTR         = mean(TTR, na.rm = TRUE),
    SD_TTR           = sd(TTR, na.rm = TRUE),
    Mean_NegIntensity= mean(NegativeIntensity, na.rm = TRUE),
    SD_NegIntensity  = sd(NegativeIntensity, na.rm = TRUE)
  )
knitr::kable(summary_stats, digits = 4,
             caption = "Descriptive Statistics by Sentiment")
```

---

## 7. Visualizations

### 7.1 TTR by Sentiment

```{r plot-ttr}
ggplot(reviews, aes(x = Sentiment, y = TTR, fill = Sentiment)) +
  geom_boxplot() +
  labs(
    title = "Type–Token Ratio by Sentiment",
    x     = "Review Sentiment",
    y     = "TTR"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### 7.2 Negative Intensity by Sentiment

```{r plot-negint}
ggplot(reviews, aes(x = Sentiment, y = NegativeIntensity, fill = Sentiment)) +
  geom_boxplot() +
  labs(
    title = "Negative Sentiment Intensity by Sentiment",
    x     = "Review Sentiment",
    y     = "Proportion of Negative Words"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## 8. Statistical Tests

### 8.1 T-test: TTR

```{r ttest-ttr}
ttr_test <- t.test(TTR ~ Sentiment, data = reviews)
ttr_test
```

> **Interpretation:**
> Negative reviews have significantly lower TTR than positive reviews (p-value < 2.2e-16).

### 8.2 T-test: Negative Intensity

```{r ttest-negint}
neg_test <- t.test(NegativeIntensity ~ Sentiment, data = reviews)
neg_test
```

> **Interpretation:**
> Negative reviews have a significantly higher proportion of negative words (p-value < 2.2e-16).

---

## 9. Regression Analysis

Create a binary indicator and run a linear model predicting TTR:

```{r regression}
reviews <- reviews %>%
  mutate(SentimentBinary = ifelse(Sentiment == "Negative", 1, 0))

reg_model <- lm(TTR ~ SentimentBinary + NegativeIntensity, data = reviews)
summary(reg_model)
```

> **Key coefficients:**
>
> * **SentimentBinary:** \~ –0.0204 (p < 2e-16)
> * **NegativeIntensity:** \~ 0.0120 (p ≈ 0.055)

---

## 10. Conclusions

* **Lexical Diversity (TTR):** Negative reviews have lower TTR than positive reviews, confirming more repetition and less variety in negative feedback.
* **Sentiment Intensity:** Negative reviews contain roughly twice the proportion of negative words compared to positive reviews.
* **Regression:** Even after accounting for the specific negative-word proportion, sentiment remains a strong predictor of lexical diversity.

**Future Directions:**

* Explore neutral (3-star) reviews as a midpoint.
* Compare across product categories.
* Analyze temporal trends in language use over time.

---

## Session Info

```{r session-info}
sessionInfo()
```

```
```
